Efficient implementation of kernel filtering

### 2007-06-29 article/Article.tex
Updated article and included some timing
darcs-hash:20070629053316-fafb9-435d466604b8e42b77976421ff1b9202efbf9049.gz

\subsection{Arbitrary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps. This is an extension to the
original paper.

@@

\subsection{Decomposition of structuring elements}
\label{sect:MMdecomp}
Morphological erosions and dilations are separable -- successive
dilation by orthogonal lines is equivalent to dilation by a rectangle
with sides equal to the line lengths. This means that any
hyper-rectangular structuring element can be constructed using lines
parallel to the image axes.

@@

\subsection{Line structuring elements}
The decompositions discussed above are important because an efficient,
recursive, implementation of erosion and dilations along lines
exists. This method was introduced in \cite{Gil1993,vanHerk1992a} and
can compute an erosion or dilation in 3 operations per pixel,
independent of structuring element length. \cite{Gil2000} recently
reduced the cost to 1.5 pixels per pixel, but the procedure is more
complicated and there are reports of no speedup in reality.

@@

A method published recently that offers improved performance
\cite{Vandroogenbroeck2005Morphological} and the ability to perform a
line opening directly (rather than using an erosion/dilation
cascade). This method, known as {\em anchor morphology} has been
implemented in the filters discussed later in the article, but
performance issues are not yet clear. The method employs histograms
and therefore needs more complex data structures when applied to
higher precision data, potentially reducing any speed advantage.

@@

Using the same strategy, the standard deviation can be efficiently computed.
It requires an accumulator image of the squared pixels values, which can be
efficiently computed at the same time than the accumulator image. The update
formulas are provided in Equations \ref{eq:accum}, \ref{eq:accSum},
\ref{eq:accum2}, \ref{eq:accSum2} and \ref{eq:accumSigma}.

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accum2D}
\caption{Recursive computation of the accumulator image - value at $(i,j)$ computed using Equation \ref{eq:accum}.\label{fig:accum}}
\end{figure}

@@

This approach to convolution has been used recently in face detection
applications \cite{Viola2004}, where rectangular filters of many
different sizes and shapes were needed.

@@

The algorithm used by a filter is set by the {\em SetAlgorithm}
method, with options of {\em BASIC}, {\em HISTO}, {\em VHGW} and {\em
ANCHOR} for the traditional ITK implementation, the moving histogram
algorithm, the van Herk, Gil, Werman algorithm and the anchor
algorithm respectively. The VHGW and ANCHOR algorithms can only be
used for Box and Poly structuring elements. VHGW is more thoroughly
tested.

@@

\subsection{Rank and mean filters -- fastRankMean}
The {\em fastRankMean} packge uses the sliding window method to
implement rank and mean filters for arbitary kernel shapes. It also
uses line versions of these kernels to provide a ``separable rank''
filter -- an fast approximation to a real rank filter as discussed in
Section \ref{sect:rank}. There are also a number of filters that
accept a mask input and return the rank or mean value of the
intersection of kernel and mask. These are experimental and haven't
been thoroughly tested.

\subsection{Rectangular convolution -- boxConvolution}
The {\em boxConvolution} uses the accumulation method described in
Section \ref{sect:meanVar} to implement rectangular mean and variance
computations. The filters implement different boundary conditions to
{\em itkMeanImageFilter}, with the mean of the kernel inside the image
being computed. The filters in this package are multithreaded and deal
with arbitary dimensions.

@@

\subsection{Rank and Mean filters}

\subsection{Box Convolution}
The relative times for optimized and standard convolution are shown in
Table \ref{tab:perfBoxConv}, and demonstrate significantly improved
performance for kernel radii greater than 2. The times indicate the
theoretical prediction that the complexity of the algorithm is
independent of kernel size.
\begin{table}[htbp]
\centering
\begin{tabular}{ccc}
\hline
Size  &  Direct &  Box \\
1  &     0.014205    &    0.0256313 \\
2  &     0.0355665   &    0.0257389 \\
3  &     0.0507895   &    0.0194342 \\
4  &     0.0745054   &    0.0152253 \\
5  &     0.124225    &    0.0149731 \\
6  &     0.187713    &    0.0149678 \\
7  &     0.27351     &    0.0152494 \\
8  &     0.411708    &    0.0155574 \\
9  &     0.518938    &    0.0156469 \\
10 &     0.682868    &    0.0158462 \\
15  &    2.02712     &    0.0165146 \\
20  &    4.46925     &    0.0176763 \\
25  &    8.43348      & 0.0182798 \\
\hline
\hline
\end{tabular}
\caption{Execution times for direct and box convolution implementations applied to the $256 \times 256$ cthead image on a ??.\label{tab:perfBoxConv}}
\end{table}

\section{Comments and Conclusions}

\section{Acknowledgments}
We thank Dr Pierre Adenot and MIMA2 confocal facilities
(\url{http://mima2.jouy.inra.fr}) for providing the 3D test image.

>>>>

\subsection{Arbitrary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps, which is an extension to the
original paper.

@@

\subsection{Decomposition of structuring elements}
\label{sect:MMdecomp}
Morphological erosions and dilations are separable -- successive
dilation by orthogonal lines is equivalent to dilation by a rectangle
with sides equal to the line lengths. This means that any
hyper-rectangular structuring element can be constructed using several
orthogonal lines, typically parallel to the axes.

@@

\subsection{Line structuring elements}
The decompositions discussed above are important because an efficient,
recursive, implementation of erosion and dilations along lines
exists. This method was introduced in \cite{Gil1993,vanHerk1992a} and
can compute an erosion or dilation in 3 operations per pixel,
independent of structuring element length. \cite{Gil2000} recently
reduced the cost to 1.5 pixels per pixel, but the procedure is more
complicated and there are reports of no speedup in practice.

@@

A method known as {\em anchor morphology} has been published recently
\cite{Vandroogenbroeck2005Morphological} that offers improved
performance and the ability to perform a line opening directly (rather
than using an erosion/dilation cascade). This method has been
implemented in the filters discussed later in the article, but
performance issues are not yet clear. The method employs histograms
and therefore needs more complex data structures when applied to
higher precision data, potentially reducing any speed advantage.

@@

Using the same strategy, the standard deviation can be efficiently
computed.  It requires an accumulator image of the squared pixels
values, which can be efficiently computed at the same time as the
original accumulator image. The update formulas are provided in
Equations \ref{eq:accum}, \ref{eq:accSum},
\ref{eq:accum2}, \ref{eq:accSum2} and \ref{eq:accumSigma}.

@@

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accum2D}
\caption{Recursive computation of the accumulator image - value at $(i,j)$ computed using Equation \ref{eq:accum}. The double hashed region is subtracted twice because it is included in both $b_{i,j-1}$ and $b_{i-1, j}$.\label{fig:accum}}
\end{figure}

@@

This approach to convolution has been used recently in face detection
applications \cite{Viola2004}, where rectangular filters of many
different sizes and shapes were needed and could be computed from the same accumulation image.

@@

The algorithm used by a filter is set by the {\em SetAlgorithm}
method, with options of {\em BASIC}, {\em HISTO}, {\em VHGW} and {\em
ANCHOR} for the traditional ITK implementation, the moving histogram
algorithm, the van Herk, Gil, Werman algorithm and the anchor
algorithm respectively. The VHGW and ANCHOR algorithms can only be
used for Box and Poly structuring elements. VHGW is more thoroughly
tested. This HISTO algorithm produces the identical results to BASIC,
and is always faster. It is the default and should be used if you need
precisely defined, non rectangular, structuring elements. If a Box
structuring element is needed then VHGW or ANCHOR should be used, as
they offer the best performance and are exact in that case. Poly
structuring elements are approximations of circles that will offer
improved performance with VHGW or ANCHOR algorithms, but the user will
need to consider whether the approximation is suitable for the
application in question.

@@

\subsection{Rank and mean filters -- fastRankMean}
The {\em fastRankMean} packge uses the sliding window method to
implement rank and mean filters for arbitary kernel shapes. It also
uses line versions of these kernels to provide a ``separable rank''
filter -- an fast approximation to a real rank filter as discussed in
Section \ref{sect:rank}. There are also a number of filters that
accept a mask input and return the rank or mean value of the
intersection of kernel and mask. These are experimental and haven't
been thoroughly tested. The filters are {\em itkRankImageFilter,
itkMovingWindowMeanImageFilter, itkFastApproxRankImageFilter,
itkFastApproxMaskRankImageFilter} and {\em
itkMaskedRankImageFilter}. itkRankImageFilter implements arbitary
shaped kernels, itkFastApproxRankImageFilter implements the separable
approximation and itkMovingWindowMeanImageFilter implements an
arbitary shaped kernel mean.

\subsection{Rectangular convolution -- boxConvolution}
The {\em boxConvolution} uses the accumulation method described in
Section \ref{sect:meanVar} to implement rectangular mean and variance
computations. The filters implement different boundary conditions to
{\em itkMeanImageFilter}, with the mean of the kernel inside the image
being computed. The filters in this package are multithreaded and deal
with arbitary dimensions. The current filters do not support the mode
of operation used in face recognition application in which an
accumulation image is repeatedly ``queried'' for mean values of
various sized kernels. However introducing such a capability is
relatively easy.

@@

\subsection{Rank and Mean filters}
Relative times for sliding window median and mean filters and
separable median filters are shown in Table \ref{tab:perfRank}. The
predicted linear complexity is observed for the sliding window
approaches (complexity of direct approach is $O(n^2)$ in 2D, reduced
to $O(n)$ by using the sliding window. The separable median exhibits a
runtime independent of kernel size, as expected.
\begin{table}[htbp]
\centering
\begin{tabular}{cccccc}
\hline
Size  &  Direct Median &  Sliding Median & Separable Median & Direct Mean & Sliding Mean\\
1     &   0.0352 & 0.0341 & 0.0378 & 0.0171 & 0.0229\\
5     &   0.409  & 0.0823  & 0.0201 & 0.24  &  0.0601 \\
10    &   1.03   & 0.131 & 0.0193 & 0.803 &  0.0569 \\
15    &   2.82   & 0.106  & 0.0192 & 2.13  &  0.0819 \\
20    &   5.85   & 0.138  & 0.0358 & 4.76  &  0.106 \\
40    &   39.7   & 0.2563  & 0.0198 & 36.6  &  0.203 \\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for median and mean filters using direct and sliding window implementations applied to the $256 \times 256$ cthead image on a 1.8GHz AMD Athlon, 512KB cache. The separable median is an approximation and therefore doesn't produce the same result as the direct or sliding algorithms.\label{tab:perfRank}}
\end{table}

\subsection{Box Convolution}
The relative times for optimized and standard convolution are shown in
Table \ref{tab:perfBoxConv}, and demonstrate significantly improved
performance for kernel radii greater than 1. The times support the
theoretical prediction that the complexity of the algorithm is
independent of kernel size.
\begin{table}[htbp]
\centering
\begin{tabular}{ccccc}
\hline
Size  &  Direct Mean &  Box Mean & Direct sigma & Box sigma\\
1    &   0.0142 & 0.0226 & 0.0176 & 0.0332 \\
2    &   0.0279 & 0.0176 & 0.0294 & 0.0258 \\
3    &   0.0425 & 0.013  & 0.0447 & 0.0193 \\
4    &   0.077  & 0.0132 & 0.0803 & 0.0195 \\
5    &   0.128  & 0.0132 & 0.129  & 0.0195 \\
6    &   0.195  & 0.0134 & 0.195  & 0.0198 \\
7    &   0.275  & 0.0136 & 0.283  & 0.0198 \\
8    &   0.392  & 0.0148 & 0.405  & 0.0204 \\
9    &   0.54   & 0.014  & 0.548  & 0.0201 \\
10   &   0.706  & 0.0139 & 0.713  & 0.0202 \\
15   &   2.12   & 0.0158 & 2.18   & 0.021 \\
20   &   4.62   & 0.0157 & 4.76   & 0.0217 \\
25   &   8.63   & 0.0164 & 8.91   & 0.0222 \\
\hline
\hline
\end{tabular}
\caption{Execution times in seconds for mean and standard deviation using direct and box convolution implementations applied to the $256 \times 256$ cthead image on a 1.8GHz AMD Athlon, 512KB cache.\label{tab:perfBoxConv}}
\end{table}
\subsection{Timing notes}
You may have noticed that the optimized filters of radius 1 seem to
regularly perform slower than their larger counterparts. We can think
of no reason for this. The algorithms that have kernel size
independent complexity should not be slower for small kernels than
large ones because the only difference occurs at borders and when the
sliding window is being created, and both situations favour small
kernels. We can only guess that there are caching issues involved.

\section{Implementation Notes}
A lot of effort has been spent implementing and optimizing these filters 
while attempting to maintain good ITK style. This section summarizes some observations
and experiences:
\subsection{Moving histograms}
The most obvious way of implementing moving histograms involves
neighborhood iterators.  However the resulting performance wasn't good
-- it appears that neighborhood iterator complexity is proportional to
radius rather than the number of active neighbors. A method using
lists of offsets combined with image Get/SetPixel methods approach was
used instead.

The original paper recommended moving the window in a zig-zag pattern, i.e horizontally 
across a the first row, down a step to the second row and then back, but this involves
accessing data in a reverse raster order which can potentially reduce cache performance.
An alternative has been implemented that always moves the window in forward raster direction,
at the cost of additional histogram copies, has been implemented and exhibits improved
performance for large images. Differences for small images weren't detectable.

\subsection{16 bit data types}
Histograms for 16 bit data types can be implemented using arrays or maps. One would
expect arrays to offer a simpler and therefore faster implementation, but our results
have been mixed. Presumably the performance depends significantly on voxel statistics.

\subsection{Line Operations}
The van Herk/Gil and Werman and anchor methods operate on image transects, potentially
at any angle. The transect data needs to be extracted from the image and the processed
transect written back, with successive parallel transects being processed. 
Existing iterators don't do this particularly efficiently, so image Get/SetPixel operations 
are being used. This approach isn't ideal either and needs to be worked on. In general
there is no need for boundary checks because the intersection between transect and image region
is calculated, so a potentially random access method without boundary checks would be ideal.

Building dimension independent code also adds complexity to this filter which can be avoided in
dimension specific implementations. The process of sweeping the transect across the image region
is much more complex (and therefore slower) when written in a dimension independent fashion.

The standard ITK threading approach isn't ideal for this style of filtering either -- a better 
option would be splitting the image based on transects instead of blocks. Obviously this
sort of change isn't practical.

\subsection{Refactoring}
The three packages do share some components, and some refactoring is
called for. This most widely used components are histograms and
sliding window infrastructure, which need to be consolidated.

\section{Comments and Conclusions}
This article has provided background theory for and implementations of
a number of important approaches to kernel based filtering. These
methods significantly reduce complexity and execution time, by some
orders of magnitude in many cases. These approaches are all well
established in the literature but weren't available in
ITK. Availability of filtering algorithms which are always fast,
irrespective of the kernel size, can make much simpler approaches to
many problems practical.
